## Basic concepts

### Installation

R is not in registry on Window. So `path/to/Rscript.exe xxx.R` will run from command line. Under the installation directory, the following subdirectories are not required for `Rscript.exe`:

- Tcl/
- tests/
- doc/
- share/
- subdir in library/
  - translations/
  - KernSmooth
  - spatial
  - survival
  - mgcv
  - rpart
  - nlme

required packages
- tools

local-installed-package, tried to delete 3, all required
- zip
- tzdb
- Rcpp

These base packages must stay in R-4.4.1/library/ and cannot be moved to user library.
`base  compiler  datasets  grDevices  graphics  grid  methods  parallel  stats  tools  utils`



### import an environment to .GlobalEnv

```{r}
# save an env object to RDS file
saveRDS(env_i, file = "env_i.RDS")

# clear global environment
rm(list = ls(envir = .GlobalEnd), envir = .GlobalEnv)

# read the saved environment
env_i <- readRDS("env_i.RDS")

# load all objects in env_i to global environment
list2env(as.list(env_i), envir = .GlobalEnv)

# delete env_i from global environment
rm("env_i)
```

### catch exit status with system()

For example, we have a test file `ttt.R`, which has a quit statement
```{r}
print("stat the test")
quit(save = "no", status = 5) # nolint
```
In another file `ttt_run.R`, which runs `ttt.R` with `system()` function:
```{r}
# pass exit status 5 to exit_status
std_out <- system("Rscript ttt.R ; echo $?", intern = TRUE)
exit_status <- std_out[length(std_out)]
print(exit_status)
```
From terminal, when we run $ Rscript ttt_run.R, it will print 5 to stdout.
